The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-08-10 paper/paper.Rmd
edits

where $\beta_{0,j}$ is the diameter-independent growth rate of species $j$; $\text{dbh}_{ij}$ is the DBH of the focal tree at the earlier census and $\beta_{\text{dbh},j}$ the slope of that species's diameter-growth relationship;  $x^{\text{comp}}_{ijk}$ is the sum of some numerical explanatory variable of all trees of competitor species $k$ and $\lambda_{jk}$ quantifies the corresponding change in growth for individuals of group $j$ from these competitors; and $\epsilon_{ij}$ is a random error term distributed $\text{Normal}(0, \sigma^2)$.

@@

We also load species information as it existed on GitHub on 2021/08/02, which includes family, genus, and species information; as well as classifications of the canopy position (canopy, canopy emergent, understory, shrub layer), drought tolerance (intolerant, resistant) and other characteristics of the species.

```{r scbi-load-sp-data}
sp_info <- read_csv("SCBI_ForestGEO_sp_ecology.csv") %>%
  select(
    sp = spcode, family, genus, species, canopy_position,
    drought_tolerance
  ) %>%
  mutate()
sp_info
```

We join this species information to our `growth_scbi` data frame.

```{r}
growth_scbi <- growth_scbi %>%
  left_join(sp_info, by = "sp") %>%
  # TODO:
  mutate(sp = as.factor(sp))
```

@@

Our model of interspecific competition relies on a spatial definition of who competitor trees are: all trees within a distance `comp_dist` of a focal tree. Here we set `comp_dist` to 7.5m, a value informed by other studies [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006], but the package could also be used to compare multiple distances and see which is best supported (see Appendix 2 \ref{mult-dist}). We use `comp_dist` and a manually constructed `sf` representation of the study region's boundary as inputs to `add_buffer_variable()` to add a `buffer` boolean variable to `growth_scbi`. All trees with `buffer` equal to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only act as competitor trees.

@@

We then identify all focal trees and their corresponding competitor trees. More specifically, identify all trees that are not part of the buffer region, have a valid `growth` measurement, and have at least one neighbor within `r comp_dist`m. We do this using `create_focal_vs_comp()`, which takes the previously detailed `comp_dist` and `id` arguments as well as the `sf` representation of the spatial cross-validation blocks and returns a new data frame `focal_vs_comp_scbi`.

@@

We observe 4 variables describing 20 competitor trees: the unique tree-stem ID, the distance to the focal tree (all $\leq$ `r comp_dist` m), the species, and the basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ for the DBH in cm from the earlier census. Saving competitor information in list-column format minimizes redundancy since we do not need to repeat information on the focal tree 20 times. We visualize the spatial distribution of these trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

Here we use basal area as the continuous predictor variable but the package is flexible to allow the user to specify any continuous predictor variable (basal area, biomass, tree height, a soil nutrient value, \dots). The package can also be used to compare predictor variables and see which best explains tree growth, see Appendix 3 \ref{compare-pred-var} for an example comparing basal area and above ground biomass. Similarly, the package can use any categorical variable as a predictor and compare between different categorical variables. For example in @allen_permutation_2020 we compare grouping individuals based on species, family, and based on trait-based groups. In Appendix 4 \ref{compare-group-var} we give another example and compare grouping individuals by species or by potential canopy position (canopy, understory, shrub layer).

@@

The resulting`comp_bayes_lm_scbi` is an object of S3 class type `comp_bayes_lm` containing the posterior values of all parameters. Furthermore, this class includes generics for three methods. First, the generic for `print()` displays the names of all prior and posterior parameters and the model formula:

@@

We intend to archive all data and source code for this manuscript on GitHub at https://github.com/rudeboybert/forestecology and on Zenodo upon acceptance. The example Smithsonian Conservation Biology Institute census data are available on GitHub at https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_main_census/data/census-csv-files and are archived on Zenodo at https://doi.org/10.5281/zenodo.2649301 [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020].

# Appendix 1: Replicate RMSE comparison

This code replicates Figure \ref{fig:scbi-simulation}: A comparison of root mean squared error of
models for standard, permuted, and spatial cross- validated error estimates.

@@

# Appendix 2: Compare different competitive distances {#mult-dist}

For all the above analyses we took the cut off distance (`comp_dist`) for two stems to compete as 7.5 m. This distance has been estimated in previous neighborhood competition studies in forests [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006]. We used 7.5 m in @allen_permutation_2020 as an average of the values estimated in other studies. But our package can be used to find which distance is best supported by the data. Here we provide an example using another section of the SCBI plot to provide an additional example of the CV block layout. To speed computation we do not consider species differences in competitive effects and treat all species as the same.

@@

Here we see that a cut off distance of approximately 6 m minimizes RMSE.

# Appendix 3: Compare predictor variables {#compare-pred-var}

@@

# Appendix 4: Compare grouping variables {#compare-group-var}

>>>>

where $\beta_{0,j}$ is the diameter-independent growth rate of species $j$; $\text{dbh}_{ij}$ is the DBH of the focal tree at the earlier census and $\beta_{\text{dbh},j}$ the slope of that species's diameter-growth relationship;  $x^{\text{comp}}_{ijk}$ is the sum of some numerical explanatory variable of all trees of competitor species $k$, and $\lambda_{jk}$ quantifies the corresponding change in growth for individuals of species $j$ from these competitors; and $\epsilon_{ij}$ is a random error term distributed $\text{Normal}(0, \sigma^2)$.

@@

We also load species information as it existed on GitHub on 2021/08/02, which includes family, genus, and species information; as well as classifications of the canopy position (canopy, canopy emergent, understory, shrub layer), drought tolerance (intolerant, resistant), and other characteristics of the species.

```{r scbi-load-sp-data}
sp_info <- read_csv("SCBI_ForestGEO_sp_ecology.csv") %>%
  select(
    sp = spcode, family, genus, species, canopy_position,
    drought_tolerance
  )
sp_info
```

We join this species information to our `growth_scbi` data frame and convert the species variable to a factor.

```{r}
growth_scbi <- growth_scbi %>%
  left_join(sp_info, by = "sp") %>%
  mutate(sp = as.factor(sp))
```

@@

Our model of interspecific competition relies on a spatial definition of who competitor trees are: all trees within a distance `comp_dist` of a focal tree. Here we set `comp_dist` to 7.5m, a value informed by other studies [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006], but the package could also be used to compare multiple distances and see which is best supported (see Appendix \ref{mult-dist}). We use `comp_dist` and a manually constructed `sf` representation of the study region's boundary as inputs to `add_buffer_variable()` to add a `buffer` boolean variable to `growth_scbi`. All trees with `buffer` equal to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only act as competitor trees.

@@

We then identify all focal trees and their corresponding competitor trees. More specifically, identify all trees that are not part of the buffer region, have a valid `growth` measurement, and have at least one neighbor within `r comp_dist`m. We do this using `create_focal_vs_comp()`, which takes the previously detailed `comp_dist` and `id` arguments, the `sf` representation of the spatial cross-validation blocks `blocks_scbi`, and a specification `comp_x_var` of the `basal_area` variable we use as the competitor explanatory variable $x^{\text{comp}}_{ijk}$ from Equation \ref{eq:model}. This function returns a new data frame `focal_vs_comp_scbi`.

@@

We observe 4 variables describing 20 competitor trees: the unique tree-stem ID, the distance to the focal tree (all $\leq$ `r comp_dist`m), the species, and the basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ for the DBH in cm from the earlier census. Saving competitor information in list-column format minimizes redundancy since we do not need to repeat information on the focal tree 20 times. We visualize the spatial distribution of these trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

Here we use basal area as the continuous predictor variable but the package is flexible to allow the user to specify any continuous predictor variable (basal area, biomass, tree height, a soil nutrient value, \dots). The package can also be used to compare predictor variables and see which best explains tree growth, see Appendix \ref{compare-pred-var} for an example comparing basal area and above ground biomass. Similarly, the package can use any categorical variable as a predictor and compare between different categorical variables. For example in @allen_permutation_2020 we compare grouping individuals based on species, family, and based on trait-based groups. In Appendix \ref{compare-group-var} we give another example and compare grouping individuals by species or by potential canopy position (canopy, understory, shrub layer).

@@

The resulting `comp_bayes_lm_scbi` is an object of S3 class type `comp_bayes_lm` containing the posterior values of all parameters. Furthermore, this class includes generics for three methods. First, the generic for `print()` displays the names of all prior and posterior parameters and the model formula:

@@

We intend to archive all data and source code for this manuscript on GitHub at https://github.com/rudeboybert/forestecology and on Zenodo upon acceptance. The example Smithsonian Conservation Biology Institute census and species information data are available on GitHub at https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/ and are archived on Zenodo at https://doi.org/10.5281/zenodo.2649301 [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020].

<div id="refs"></div>

\appendix

# Appendix: Replicate RMSE comparison

This code replicates Figure \ref{fig:scbi-simulation}: A comparison of root mean squared error of
models for standard, permuted, and spatial cross-validated error estimates.

@@

# Appendix: Compare different competitive distances {#mult-dist}

For all the above analyses we took the cut off distance (`comp_dist`) for two stems to compete as `r comp_dist`m. This distance has been estimated in previous neighborhood competition studies in forests [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006]. We used `r comp_dist`m in @allen_permutation_2020 as an average of the values estimated in other studies. But our package can be used to find which distance is best supported by the data. Here we provide an example using another section of the SCBI plot to provide an additional example of the CV block layout. To speed computation we do not consider species differences in competitive effects and treat all species as the same.

@@

Here we see that a cut off distance of approximately 6m minimizes the RMSE.

# Appendix: Compare predictor variables {#compare-pred-var}

@@

# Appendix: Compare grouping variables {#compare-group-var}
